---
title: "Data Documentation"
output: html_document
date: "2023-12-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data Documentation

```{r}
library(tidyverse)

demographic_data <- read_delim("acs2017_county_data.csv")
us_census <- read_delim("president_county.csv")

final_dataset <- read_delim("voter_demographic.csv")
names(final_dataset)
```

We found the ACS dataset, acs2017_county_data.csv, at the following link: 
https://www.kaggle.com/datasets/muonneutrino/us-census-demographic-data. The
data was collected by mounneutrino.

We found the election dataset, president_county.csv, at the following link: 
https://www.kaggle.com/datasets/unanimad/us-election-2020?select=president_county.csv. The
data was collected by Raphael Fontes.

We merged the two datasets to create the dataset: voter_demographic.csv.

Within our merged dataset, the rows represent data about a US county (ex. King County, WA).
There are 4633 observations and 15 variables in our final dataset. The variables State, County,
TotalVotes, and total_votes_by_state have no missing values, while the variables Men, TotalPop,
Women, Hispanic, White, Black, Native, Asian, Pacific, VotingPop, and non_voters have 1589 
missing values. This can be explained by the data quality issues we experienced, where missing
values were caused by problems with merging the datasets when there wasn't demographic
information present about a County. 

The variables we will be using for this report are:

* State: represents the name of the state where the data was collected.
* County: represents the name of the county where the data was collected.
* TotalVotes: represents the total amount of votes made in a county.
* Men: represents the total number of men in the county.
* TotalPop: represents the total number of people in the county.
* Women: represents the total number of women in the county.
* Hispanic: represents the total percentage of Hispanic/Latino people in the county.
* White: represents the total percentage of white people in the county.
* Black: represents the total percentage of black people in the county.
* Native: represents the total percentage of Native Americans in the county.
* Asian: represents the total percentage of Asians in the county.
* Pacific: represents the total percentage of Pacific Islanders in the county.
* VotingPop: represents the total number of people eligble to vote in the county.
* total_votes_by_state: represents the total number of votes in each state.
* non_voters: represents the total number of eligible voters that didn't vote in 2020.