---
title: "Final Project"
output: html_document
date: "2023-12-01"
---

```{r}
library(tidyverse)

demographic_data <- read_delim("acs2017_county_data.csv")
us_census <- read_delim("president_county.csv")
```

Cleaning and Merging Data:
```{r}
demographic_data <- demographic_data %>% 
  select(State, County, Men,TotalPop, Women, Hispanic, White, Black, Native, Asian, Pacific, "VotingPop" = VotingAgeCitizen)


us_census <- us_census %>% 
  select("State" = state, "County" = county, "Total_votes" = total_votes)

census_demographic_merge <- left_join(us_census,demographic_data, by= c("State","County"))
```

**Data Documentation**

The first dataset we used was the acs2017_county_data.csv file. We found this dataset at the
following link: https://www.kaggle.com/datasets/muonneutrino/us-census-demographic-data. The
data was collected by mounneutrino and taken from the 2015 American Community Survey 5-year
estimates.

The second dataset we used was the president_county.csv file. We found this dataset at the
following link: 
https://www.kaggle.com/datasets/unanimad/us-election-2020?select=president_county.csv. The
data was collected by Raphael Fontes and taken from 2020 US Election data.

We cleaned and merged these two datasets to create the dataset: census_demographic_merge.
```{r}
head(census_demographic_merge,3)
```
 This dataset has `r nrow(census_demographic_merge)` rows,`r ncol(census_demographic_merge)`,
 columns, and the following number of missing values for each column:
```{r}
 colNA <- colSums(is.na(census_demographic_merge))
colNA
```

The variables we will be using for this report are: `r names(census_demographic_merge)`.

* State: represents the name of the state where the data was collected.
* County: represents the name of the county where the data was collected.
* Total_votes: represents the total amount of votes made in a county.
* Men: represents the total number of men in the county.
* TotalPop: represents the total number of people in the county.
* Women: represents the total number of women in the county.
* Hispanic: represents the total percentage of Hispanic/Latino people in the county.
* White: represents the total percentage of white people in the county.
* Black: represents the total percentage of black people in the county.
* Native: represents the total percentage of Native Americans in the county.
* Asian: represents the total percentage of Asians in the county.
* Pacific: represents the total percentage of Pacific Islanders in the county.
* VotingPop: represents the total number of people eligble to vote in the county.

Question 1:






Question 3: Identifying Male vs. Female dominant voter populations (by county and state)

Here, I am calculating the number of men and women that voted in each county in the 2020 election.
```{r}
gender <- census_demographic_merge %>% 
  select(State, Total_votes, County, Men, Women, VotingPop, TotalPop)

gender <- gender %>% 
  group_by(County) %>% 
  mutate(femaleVoterPercentage = Women / TotalPop) %>% 
  mutate(femaleVoters = round(femaleVoterPercentage * Total_votes)) %>% 
  mutate(maleVoters = Total_votes - femaleVoters)
```

Now that I have found the number of voters of each gender, I can find what are the counties
and states with the highest populations of male and female voters.

Below are the 3 Counties with the largest Female Voter Populations:
```{r}
gender %>% 
  arrange(desc(femaleVoterPercentage)) %>% 
  select(State,County, femaleVoters, maleVoters) %>% 
  head(3)
```

Below are the 3 States with the largest Female Voter Populations:
```{r}
# finds the states with the largest number of total female voters.
stateFemaleVoters <- gender %>% 
  group_by(State) %>% 
  summarize(totalFemaleVoters = sum(femaleVoters, na.rm = TRUE)) %>% 
  arrange(desc(totalFemaleVoters))
head(stateFemaleVoters,3)


# find the states with the largest the number of female majority voting counties
test <- gender %>% 
  group_by(State) %>% 
  filter(femaleVoterPercentage > 0.5) %>% 
  summarize(TotalFemaleMajCounties = sum(femaleVoters, na.rm = TRUE)) %>% 
  arrange(desc(TotalFemaleMajCounties))
head(test, 3)

# find the states with the largest difference between male and female voters
genderDiff <- gender %>% 
  group_by(State) %>% 
  summarize(genderDiff = sum(femaleVoters) - sum(maleVoters)) %>% 
  arrange(desc(genderDiff)) %>% 
  head(10)
  
  ggplot(genderDiff,aes(State,genderDiff, fill = State)) + geom_col()
```

