---
title: "Final Project"
output: html_document
date: "2023-12-01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

##    Context

We chose this topic because the relationship between demographics and voting patterns offers
valuable insight into understanding the influence of these trends on political outcomes in 
the United States. Historically, the United States has witnessed shifts in racial, ethnic and
gender compositions. These shifts have the potential to change the political landscape which
directly influences voters. They’re especially significant for the 2020 presidential election
due to heightened political tension, polarization, and its impact on the global political
climate.

Furthermore, politicians have utilized demographic data to shape their political campaigns.
These political strategies require an understanding of this data to influence voting patterns
and resonate with a diverse voter population. Also in 2020, social movements like “Black Lives
Matter,” had a significant influence on voter engagement among different demographic groups
which politicians had to analyze and adapt to. For example, research has shown that younger
voters tend to have lower voter turnouts but heavily lean towards progressive politics while
older voters show higher participation, but their political opinions can vary.

Race and gender are pivotal factors that affect this data. Historically, groups part of these
communities were confronted with barriers to voting and are a critical factor in analyzing 
rising political engagement. Black and Hispanic communities tend to lean towards the Democratic
party with rising voter turnout. Women’s voter turnout has been consistently higher than men
for the last few decades and they’re voting priorities often differ from their counterparts.

The 2020 US elections provided an opportunity to better understand these demographic trends
and voting patterns. This project is designed to uncover broader patterns of barriers to
voting through analyzing the relationships between race, gender, and voter turnout. We plan
to identify three states or counties with the highest number of nonvoters and the top ten 
states with the lowest overall voter turnout. This will show where voter engagement is 
particularly low and uncover challenges specific to the demographic. We will be merging voter
data from each county to determine the total number of voters by state, and this will provide
a macro perspective on voter engagement. We will be answering these research questions 
specifically: Does race impact the number of people voting in each area? What percentage of
each demographic is voting? Is there a relationship between race and number of votes? Does 
gender affect the number of voters in each state?

This topic of the 2020 elections is interesting because it provides a new window into American
politics. It had unprecedented voter turnout and intense political engagement from various 
groups which offered an opportunity to study how social dynamics and movements affect voter
turnout. Understanding how demographics affect voter turnout is especially important for 
political scientists, sociologists, policy makers and the public. This knowledge is key to 
shaping effective political strategies and making policies that resonate with the voters. It
allows them to understand what motivates the population to participate in voting and what 
doesn’t. For the public, this information will help them make informed decisions on how 
their demographic identity influences their political decisions. Reading our report will 
provide these groups with a detailed and nuanced analysis of the 2020 election data. They can
use this report to gain a better understanding of the complexities of demographic influences
and provide an analysis that goes beyond surface level observations. 

##    Questions

### 1. Which states/counties have the highest number of non-voters?

### 2. What is the race distribution of voters in the US census?

### 3. Does gender affect the number of voters in each state?

# Data

##    Why These Datasets

We chose these datasets because they provided comprehensive and detailed information from
both county and census levels from across the United States. Both datasets allow for a deeper
insight into understanding the correlation between demographics and voter behaviors and are
crucial for political analysis. Data from the American Community survey (ACS) provided 
demographic information regarding age, race and employment. The ACS data is derived from the
American Factfinder website and is used by many across the nation as the primary source for 
detailed demographic information from the entire country. The data held on the website was 
collected by the US Census Bureau, a widely known and trusted source in the United States. 
Since it is a product of a federal agency, it is not subject to any copyright and is fully 
accessible to the public. The 2020 presidential election data provided information into voting
patterns. It is not explicitly stated who collected the data, but it is a compilation of 
official results from state and county offices which makes it a reliable resource.  


##    Data Information

Both datasets were found on Kaggle which is a widely known website in the data science
community that allows for data sharing and analysis. Kaggle’s policies make it clear that
any dataset found on their platform should be used for noncommercial use only and not on
behalf of any third party. The population of the dataset from the ACS is the United States
including Puerto Rico. The sample is a subset of this population to make educated estimates
about the population. The ACS did not survey every household in the United States. The 
population of the dataset from the 2020 US Presidential election is the total number of 
eligible voters in each county, regardless of whether they voted, across the United States
in the 2020 Presidential election. There is no sample for this dataset because it is a 
comprehensive record of the whole population. Both of these datasets provide a detailed 
view of the United States population which is essential for understanding trends in demographics.

Both datasets were found on Kaggle which is a widely known website in the data science 
community that allows for data sharing and analysis. Kaggle’s policies make it clear that
any dataset found on their platform should be used for noncommercial use only and not on
behalf of any third party. The population of the dataset from the ACS is the United States
including Puerto Rico. The sample is a subset of this population to make educated estimates
about the population. The ACS did not survey every household in the United States. The 
population of the dataset from the 2020 US Presidential election is the total number of 
eligible voters in each county, regardless of whether they voted, across the United States
in the 2020 Presidential election. There is no sample for this dataset because it is a 
comprehensive record of the whole population. Both of these datasets provide a detailed view
of the United States population which is essential for understanding trends in demographics.

We found the ACS dataset, acs2017_county_data.csv, at the following link: 
https://www.kaggle.com/datasets/muonneutrino/us-census-demographic-data. The
data was collected by mounneutrino.

We found the electio dataset, president_county.csv, at the following link: 
https://www.kaggle.com/datasets/unanimad/us-election-2020?select=president_county.csv. The
data was collected by Raphael Fontes.

##    Data Quality Issues

While analyzing the data, there were some challenges with missing values and inconsistencies
in data reporting between counties. When merging the two datasets, there were some discrepancies
in how the demographic categories were defined. To manage these issues, we used data cleaning
techniques and made sure to carefully interpret the data keeping these limitations in mind.

##    Final Dataset

We cleaned and merged these two datasets to create the dataset: census_demographic_merge.

###   Data Description

The rows represent data about a US county (ex. King County, WA). There are 4633 observations
and 15 variables in our final dataset. The variables State, County, TotalVotes, and
total_votes_by_state have no missing values, while the variables Men, TotalPop, Women,
Hispanic, White, Black, Native, Asian, Pacific, VotingPop, and non_voters have 1589 missing
values. This can be explained by the data quality issues we experienced, where missing
values were caused by problems with merging the datasets when there wasn't demographic
information present about a County. 

The variables we will be using for this report are:

* State: represents the name of the state where the data was collected.
* County: represents the name of the county where the data was collected.
* TotalVotes: represents the total amount of votes made in a county.
* Men: represents the total number of men in the county.
* TotalPop: represents the total number of people in the county.
* Women: represents the total number of women in the county.
* Hispanic: represents the total percentage of Hispanic/Latino people in the county.
* White: represents the total percentage of white people in the county.
* Black: represents the total percentage of black people in the county.
* Native: represents the total percentage of Native Americans in the county.
* Asian: represents the total percentage of Asians in the county.
* Pacific: represents the total percentage of Pacific Islanders in the county.
* VotingPop: represents the total number of people eligble to vote in the county.
* total_votes_by_state: represents the total number of votes in each state.
* non_voters: represents the total number of eligible voters that didn't vote in 2020.

# Methods

### Loading The Data
```{r}
library(tidyverse)

demographic_data <- read_delim("acs2017_county_data.csv")
us_census <- read_delim("president_county.csv")
```
### Cleaning and Merging Data
```{r}
demographic_data <- demographic_data %>% 
  select(State, County, Men,TotalPop, Women, Hispanic, White, Black, Native, Asian, Pacific, "VotingPop" = VotingAgeCitizen)


us_census <- us_census %>% 
  select("State" = state, "County" = county, "TotalVotes" = total_votes)

census_demographic_merge <- left_join(us_census,demographic_data, by= c("State","County"))

head(census_demographic_merge,3)
```

We included the variables State, County, Men,TotalPop, Women, Hispanic, White, Black, Native,
Asian, Pacific, and VotingPop from the American Community Survey dataset. We included the 
variables State, County, and TotalVotes from the election dataset. We used the County Variable
as the main variable for merging the two datasets. We renamed variable names for convenience 
and understanding as well as to enable proper merging. 

## Question 1. Which State has the highest demographic of each voting population?

First, akdjfasdklfjsdalkfjlks;df;jlksaljk;dsafasjlk;

```{r}
census_demographic_merge <- census_demographic_merge %>%
  group_by(State) %>%
  mutate(total_votes_by_state = sum(TotalVotes))

head(census_demographic_merge,5)

```

```{r}

census_demographic_new <- census_demographic_merge %>%
  pivot_longer(cols = c(Hispanic, White, Black, Native, Pacific, Asian),
               names_to = "Ethnicity",
               values_to = "Votes") %>%
  group_by(Ethnicity) %>%
  summarize(TotalVotes = sum(Votes, na.rm = TRUE))

# Plotting
ggplot(census_demographic_new, aes(x = Ethnicity, y = TotalVotes, fill = Ethnicity)) +
  geom_bar(stat = "identity") +
  labs(
    x = "Ethnicity",
    y = "Total Number of Votes",
    title = "Total Number of Votes by Ethnicity Across States"
  ) +
  theme_minimal()

```

## Question 2: Find the Top three counties in the country that have the highest ratio of non-voters to eligible voting population.

```{r}
#make a new colums that calculates the number of non voters

# Assuming your data frame is named 'census_demographic_merge'
census_demographic_merge <- census_demographic_merge %>%
  mutate(non_voters = VotingPop - TotalVotes)

# Print or view the updated data frame
print(census_demographic_merge)

```

```{r}
top3_non_voters_counties <- census_demographic_merge %>%
  group_by(County) %>%
  summarize(total_non_voters = sum(non_voters, na.rm = TRUE),
            total_VotingPop = sum(VotingPop, na.rm = TRUE)) %>%
  arrange(desc(total_non_voters)) %>%
  slice_head(n = 3)

# Plotting
ggplot(top3_non_voters_counties, aes(x = reorder(County, -total_non_voters), 
                                     fill = factor(c("VotingPop", "non_voters")))) +
  geom_bar(aes(y = total_VotingPop), stat = "identity", position = "stack", fill = "pink") +
  geom_bar(aes(y = total_non_voters), stat = "identity", position = "stack", fill = "purple") +
  labs(
    x = "County",
    y = "Count",
    title = "Top 3 Counties with the Highest Number of Non-Voters and VotingPop"
  ) +
  scale_fill_manual(values = c("VotingPop" = "pink", "non_voters" = "purple")) +
  theme_minimal()
```

### Plotting the ethnicity breakdown of the top 3 non-voting counties

```{r}
#for Los Angeles County

# Filter data for Los Angeles County
la_county_data <- census_demographic_merge %>%
  filter(County == "Los Angeles County")

# Pivot the data into long format for easier plotting
la_county_data_long <- la_county_data %>%
  pivot_longer(cols = c(Hispanic, White, Black, Native, Pacific, Asian),
               names_to = "Ethnicity",
               values_to = "Votes")

# Plotting
ggplot(la_county_data_long, aes(x = Ethnicity, y = Votes, fill = Ethnicity)) +
  geom_bar(stat = "identity") +
  labs(
    x = "Ethnicity",
    y = "Number of Votes",
    title = "Ethnicity Breakdown in Los Angeles County"
  ) +
  theme_minimal()

#For Cook County
cook_county_data <- census_demographic_merge %>%
  filter(County == "Cook County")
cook_county_data_long <- cook_county_data %>%
  pivot_longer(cols = c(Hispanic, White, Black, Native, Pacific, Asian),
               names_to = "Ethnicity",
               values_to = "Votes")
ggplot(cook_county_data_long, aes(x = Ethnicity, y = Votes, fill = Ethnicity)) +
  geom_bar(stat = "identity") +
  labs(
    x = "Ethnicity",
    y = "Number of Votes",
    title = "Ethnicity Breakdown in Cook County"
  ) +
  theme_minimal()

#For Harris County

# Filter data for Harris County
harris_county_data <- census_demographic_merge %>%
  filter(County == "Harris County")

# Pivot the data into long format for easier plotting
harris_county_data_long <- harris_county_data %>%
  pivot_longer(cols = c(Hispanic, White, Black, Native, Pacific, Asian),
               names_to = "Ethnicity",
               values_to = "Votes")

# Plotting
ggplot(harris_county_data_long, aes(x = Ethnicity, y = Votes, fill = Ethnicity)) +
  geom_bar(stat = "identity") +
  labs(
    x = "Ethnicity",
    y = "Number of Votes",
    title = "Ethnicity Breakdown in Harris County"
  ) +
  theme_minimal()

```

# Question 3: Identifying Male vs. Female dominant voter populations (by county and state)

### Part 1
Here, I am calculating the number of men and women that voted in each county in the 2020 election.
I selected the variables State, TotalVotes, County, Men, Women, VotingPop, and TotalPop, into a new
data frame to first derive the total number of female and male voters. To do this, I grouped by County
and calculated the female population percentage by taking the number of women divided by the total 
population. Then, I multiplied that by the number of total votes to find the proportion of votes that
were women in each county. I rounded this value, and finally calculated by number of male voters. I
didn't use more complex statistical methods, nor did I remove or replace missing values. Missing values
did not affect my output. 

```{r}
gender <- census_demographic_merge %>% 
  select(State, TotalVotes, County, Men, Women, VotingPop, TotalPop)

gender <- gender %>% 
  group_by(County) %>% 
  mutate(femaleVoterPercentage = Women / TotalPop) %>% 
  mutate(femaleVoters = round(femaleVoterPercentage * TotalVotes)) %>% 
  mutate(maleVoters = TotalVotes - femaleVoters)
```

### Part 2
Now that I found the number of voters of each gender, I could find what the counties and states
are with the largest difference between male and female voters. To find these counties, I 
arranged the data in descending order based on the percentage of the population that is 
female, and selected the variables State, County, femaleVoters and MaleVoters, and returned
the top 3 counties. I didn't use more complex statistical methods, nor did I remove or 
replace missing values. Missing values did not affect my output. 

Below are the 3 Counties with the largest Female Voter Percentage:
```{r}
gender %>% 
  arrange(desc(femaleVoterPercentage)) %>% 
  select(State, County, femaleVoters, maleVoters, femaleVoterPercentage) %>% 
  head(3)
```

### Result
The table shows us that the counties with the largest difference between female and male voters
are Norton City in Virginia, Pulaski County in Georgia, and Owsley County in Kentucky. This shows
us that these Counties have some of the highest voter turnouts from women. One thing to note is
that they have relatively small populations, likely causing greater variation and a higher 
percentage. 

### Part 3
To find the states with the largest female voter populations, I grouped the dataset by state,
summarized the total number of female voters in each state, arranged the states in descending
order based on the total number of female voters, and displayed the top 3 states with the
largest female voter populations.

Below are the 3 states with the largest Female Voter Populations:
```{r}
stateFemaleVoters <- gender %>% 
  group_by(State) %>% 
  summarize(totalFemaleVoters = sum(femaleVoters, na.rm = TRUE)) %>% 
  arrange(desc(totalFemaleVoters))
head(stateFemaleVoters,3)

totalFemalePop <- stateFemaleVoters %>% 
  summarize(overallTotal = sum(totalFemaleVoters))
```

### Result
Here, the data tells us that California, Texas and Florida have the largest female voter 
populations. These three states have a combined total of over 20 million women, a 
significant portion of the population. Additionally, considering that the total number of
female voters is `r totalFemalePop$overallTotal`, these three states make up a considerable
percent of the total female voting population.

### Part 4
To find the states with the largest female majority voting counties, I identified states 
where the percentage of female voters was greater than 50%, summarized the total number of
counties in each state where females are the majority voters and displayed the top 3 states
with the largest number of female majority voting counties.

Below are the 3 states with the largest Female Majority Voting Counties:
```{r}
test <- gender %>% 
  group_by(State) %>% 
  filter(femaleVoterPercentage > 0.5) %>% 
  summarize(TotalFemaleMajCounties = sum(femaleVoters, na.rm = TRUE)) %>% 
  arrange(desc(TotalFemaleMajCounties))
head(test, 3)
```

Here, the data shows that the states California, Florida, and Texas have the most counties
where there are more women than men. This is interesting because it displays that women
are being represented in politics and that their voice is being heard, as there are many
counties where they make up the majority of the county population. In local elections for
mayors, representatives and other political offices, women will and are able to ensure
their voice in the political process.

### Part 5
To find the States with the largest difference in male and female voters, where they had
more women than men, I calculated the difference between the total number of female voters
and male voters in each state, arranged the states based on the gender difference, and 
displayed the top 10 states with the largest gender differences in a bar plot.


Below are the 10 States with the largest difference of male and female voters:
```{r}
genderDiff <- gender %>% 
  group_by(State) %>% 
  summarize(genderDiff = sum(femaleVoters) - sum(maleVoters)) %>% 
  arrange(desc(genderDiff)) %>% 
  head(10)
  
  ggplot(genderDiff,aes(State,genderDiff, fill = State)) + geom_col()
```

This data takes the previous part a step further, and goes past female majority counties to 
look at female majority states, where we see some of the most populated states, like California
New York and Florida. This plot shows that women have a signficant voice in the political
process, as in some of the biggest states in the country, they have tens and even hundreds of
thousands more votes than men. 

# The Final Cleaned Dataset

After completing all of our data manipulations, here we write the dataset as a csv file
to download. We name the file voter_demographic.csv. 
```{r}
write_csv(census_demographic_merge, file = "voter_demographic.csv")
```

# Findings

##    What Do the Results Tell Us About Our Question/Story

##    Limitations

##    Future Work

# Summary
